---
title: "Integrated analysis with Seurat of atherosclerotic plaques in the context of Myeloid PHD2 deficiency"
author: "Javier Perales-Patón - javier.perales@bioquant.uni-heidelberg.de"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load libraries
```{r}
suppressPackageStartupMessages(require(Seurat))
suppressPackageStartupMessages(require(GSEABase))
suppressPackageStartupMessages(require(dplyr))
suppressPackageStartupMessages(require(clustree))
suppressPackageStartupMessages(require(cowplot))
suppressPackageStartupMessages(require(genesorteR))
suppressPackageStartupMessages(require(optparse))
# Get some functions for Seurat analysis
source("../src/seurat_fx.R")
set.seed(1234)
```

## Setting-up environment
### Input parameters
```{r}
# Define input parameters
option_list = list(
  make_option(c("--CASE_INPUT"), action="store", default="../data/sc/CK64/filtered_feature_bc_matrix/", type='character',
              help="cellranger count folder that contains the output matrices"),
  make_option(c("--CASE_SNAME"), action="store", default="PHD2cKO", type='character',
              help="Sample name"),
  make_option(c("--CONTROL_INPUT"), action="store", default="../data/sc/CK62/filtered_feature_bc_matrix/", type='character',
              help="cellranger count folder that contains the output matrices"),
  make_option(c("--CONTROL_SNAME"), action="store", default="WT", type='character',
              help="Sample name"),
  make_option(c("--OUTDIR"), action="store", default="./output/01_integration/", type='character',
              help="Output directory"),
  make_option(c("--NPC"), action="store", default=50, type='numeric',
              help="number of principal components to calculate."),
  make_option(c("--NPC_ANCHOR"), action="store", default=20, type='numeric',
              help="number of principal components to consider for the anchoring."),
  make_option(c("--NPC_CLUSTERING"), action="store", default=25, type='numeric',
              help="Number of principal components to consider for the cell clustering."),
  make_option(c("--RES"), action="store", default=0.1, type='numeric',
              help="Resolution for cell clustering.")
)

# Parse the parameters
opt = parse_args(OptionParser(option_list=option_list))

# Cat the input parameters
cat("[INFO] Input parameters\n", file=stdout())
for(user_input in names(opt)) {
  if(user_input=="help") next;
  cat(paste0("[INFO] ",user_input," => ",opt[[user_input]],"\n"),file = stdout())
  assign(user_input,opt[[user_input]])
}
```

### Output directory
```{r}
if(!dir.exists(OUTDIR)) dir.create(OUTDIR);
```

## Load data
Read samples and estimate percentage of mitochondrial gene expression

```{r}
CASE <- getSeuratObject(path = CASE_INPUT, project_name = CASE_SNAME, mt.pattern = "^mt-", min.cells = 5, min.features = 500)
CASE$stim <- CASE_SNAME
CASE <- subset(CASE, subset = nFeature_RNA > 500 & nFeature_RNA < 4000 & percent.mt < 7.5)
CASE <- NormalizeData(CASE, verbose = FALSE)
CASE <- FindVariableFeatures(CASE, selection.method = "vst", nfeatures = 2000)
  
CTRL <- getSeuratObject(path = CONTROL_INPUT, project_name = CONTROL_SNAME, mt.pattern = "^mt-", min.cells = 5, min.features = 500)
CTRL$stim <- CONTROL_SNAME
CTRL <- subset(CTRL, subset = nFeature_RNA > 500 & nFeature_RNA < 4000 & percent.mt < 7.5)
CTRL <- NormalizeData(CTRL, verbose = FALSE)
CTRL <- FindVariableFeatures(CTRL, selection.method = "vst", nfeatures = 2000)
```

## Perform integration
The integration consists on Find anchors between the two samples and perform CCA to integrate both.

```{r}
anchors <- FindIntegrationAnchors(object.list = list(CTRL, CASE), dims = 1:NPC_ANCHOR)
S <- IntegrateData(anchorset = anchors, dims = 1:20)
# Relevel wildtype as ref level
S$stim <- factor(S$stim, levels=c("WT","PHD2cKO"))

# Get some RAM space: we don't need this anymore
rm(CASE, CTRL)
```
We define the Integrated data for cell clustering and assignment
```{r}
DefaultAssay(S) <- "integrated"
```

## Cell clustering and dimensionality reduction
We use the default parameters from Seurat to do so.

First scale the data and get principal components
```{r}
S <- ScaleData(S, verbose = FALSE)
S <- RunPCA(S, npcs = NPC, verbose = FALSE)
```

Show the elbowplot for the PCs
```{r}
ElbowPlot(S,ndims = NPC) + geom_vline(xintercept = NPC_CLUSTERING, col="red")
```

Perform SNN clustering
```{r}
S <- FindNeighbors(S, reduction = "pca", dims = 1:NPC_CLUSTERING)
S <- FindClusters(S, resolution = seq(from=0.1, to=1.5, by=0.1))
```

```{r clustree, fig.width=14, fig.height=7, dpi=300, fig.path="./output/01_integration/", dev=c('png','tiff')}
clustree(S, prefix = "integrated_snn_res.")
```

Decide the final resolution: `r RES`
```{r}
S <- FindClusters(S, resolution = RES)
```

Finally, we use UMAP to get 2 reduced dimensions for visualization
```{r}
S <- RunUMAP(S, reduction = "pca", dims = 1:NPC_CLUSTERING)
```

```{r umap_plot, dpi=300, fig.path="./output/01_integration/", dev=c('png','tiff')}
DimPlot(S, reduction = "umap", group.by = "seurat_clusters", label = TRUE, label.size = 14)
```

```{r umap_condition, dpi=300, fig.path="./output/01_integration/", dev=c('png','tiff')}
DimPlot(S, reduction = "umap", group.by = "stim",cols = c("grey","red"))
```

### Extract cluster markers
The assay is switched to RNA

```{r}
DefaultAssay(S) <- "RNA"
# cl1.markers <- FindConservedMarkers(S, ident.1 = 1, grouping.var = "stim", verbose = FALSE)
# head(cl1.markers)
# S <- NormalizeData(S, verbose = FALSE) # I am not sure if I have to normalize again
S <- Seurat_scaledata(S)
```

```{r wilcox_test}
MARKERS.OUTDIR <- paste0(OUTDIR,"/markers")
if(!dir.exists(MARKERS.OUTDIR)) dir.create(MARKERS.OUTDIR);

up <- setNames(vector("list",length=length(levels(S))), levels(S))
for(idx in names(up)) {
  up.idx <- FindConservedMarkers(S,ident.1 = idx,ident.2 = setdiff(levels(S), idx),
				 grouping.var = "stim",only.pos=T)
  cols_names <- colnames(up.idx)

  # Add two extra cols
  up.idx$cluster <- idx
  up.idx$gene <- rownames(up.idx)

  write.table(up.idx[,c("cluster", "gene", cols_names)],
              file = paste0(MARKERS.OUTDIR,"/cluster",idx,".tsv"),
              sep="\t",col.names = TRUE, row.names = FALSE, quote=FALSE
  )

  up[[idx]] <- up.idx
}
```

```{r genesorteR}
sg <- sortGenes(S@assays$RNA@data, Idents(S))

#define a small set of markers
mm = getMarkers(sg, quant = 0.975)

#cluster genes and make a heatmap
pp = plotMarkerHeat(sg$inputMat, sg$inputClass, mm$markers, clusterGenes=TRUE, outs = TRUE)

# #the top 25 genes for each cluster by specificity scores
# top_markers = apply(sg$specScore, 2, function(x) names(head(sort(x, decreasing = TRUE), n = 25)))
# write.table(top_markers, paste0(MARKERS.OUTDIR,"/top25markers.tsv"),
# 	    sep="\t",row.names = FALSE,col.names = TRUE,quote=FALSE)
# 

# Save the tables of genesorteR stats
write.table(sg$condGeneProb, file=paste0(OUTDIR,"/condGeneProb.tsv"),sep="\t", row.names=TRUE, col.names = NA, quote=FALSE)
write.table(sg$specScore, file=paste0(OUTDIR,"/specScore.tsv"),sep="\t", row.names=TRUE, col.names = NA, quote=FALSE)
# Save the object
saveRDS(sg,file = paste0(OUTDIR,"/sg.rds"))
```

## Align clustering with previous studies in atherosclerosis and cell type markers
### Internal
```{r heatmap_specscore_top10, fig.width=14, fig.height=9}
GSC_markers <- GeneSetCollection(sapply(colnames(sg$specScore),function(idx) GeneSet(names(head(sort(sg$specScore[,idx], decreasing = TRUE), n = 10)),
                                                                                     setName=idx, shortDescription="genesorteR")))

DoHeatmap3(SeuratObject = S, GSC = GSC_markers,
           assay = "RNA", res=NULL, show_hr = FALSE)
```

### External (literature)
```{r heatmap_wirka, fig.width=14,fig.height=14, dpi=300, fig.path="./output/01_integration/", dev=c('png','tiff')}
DoHeatmap3(SeuratObject = S, GSC = getGmt("../data/markers/Wirka_etal2019.gmt"), 
	   assay = "RNA", res=NULL, show_hr=FALSE)
```

```{r heatmap_sluimer, fig.width=14, fig.height=10, dpi=300, fig.path="./output/01_integration/", dev=c('png','tiff')}
DoHeatmap3(SeuratObject = S, GSC = getGmt("../data/markers/markers.gmt"), 
	   assay = "RNA", res=NULL, show_hr=FALSE)
```

```{r heatmap_cochain, fig.width=14, fig.height=14, dpi=300, fig.path="./output/01_integration/", dev=c('png','tiff')}
DoHeatmap3(SeuratObject = S, GSC = getGmt("../data/markers/Cochain_etal2019.gmt"), 
	   assay = "RNA", res=NULL, show_hr=FALSE)
```
```{r heatmap_gu, fig.width=14, fig.height=10, dpi=300, fig.path="./output/01_integration/", dev=c('png','tiff')}
DoHeatmap3(SeuratObject = S, GSC = getGmt("../data/markers/Gu_etal2019.gmt"), 
	   assay = "RNA", res=NULL, show_hr=FALSE)
```

## Density of clusters per conditions

```{r}
cl.cnt<-tapply(S$seurat_clusters,S@meta.data$stim,table)
# sapply(names(cl.cnt), function(idx) sum(S$stim==idx))
cl.prop <- cl.cnt

for(idx in names(cl.prop)) {
  cl.prop[[idx]] <- (cl.prop[[idx]]/sum(S$stim==idx)*100)
}

DATA <- data.frame(cluster=factor(levels(S), levels(S)),
                   S1=as.numeric(cl.prop[[1]]),
                   S2=as.numeric(cl.prop[[2]]))
```

```{r cluster_density, fig.width=7, fig.height=9, dpi=300, fig.path="./output/01_integration/", dev=c('png','tiff')}
# Source: https://stackoverflow.com/questions/18265941/two-horizontal-bar-charts-with-shared-axis-in-ggplot2-similar-to-population-pyr
library(grid)
g.mid<-ggplot(DATA,aes(x=1,y=cluster))+geom_text(aes(label=as.character(DATA$cluster)), size=14)+
  geom_segment(aes(x=0.94,xend=0.96,yend=cluster))+
  geom_segment(aes(x=1.04,xend=1.065,yend=cluster))+
  ggtitle("Cluster")+
  ylab(NULL)+
  scale_x_continuous(expand=c(0,0),limits=c(0.94,1.065))+
  theme(plot.title = element_text(size=14),
        panel.background = element_rect(fill="white"),
        axis.title=element_blank(),
        panel.grid=element_blank(),
        axis.text.y=element_blank(),
        axis.ticks.y=element_blank(),
        axis.text.x=element_text(color=NA, size=14),
        axis.ticks.x=element_line(color=NA),
        plot.margin = unit(c(1,-1,1,-1), "mm"))

g1 <- ggplot(data = DATA, aes(x = cluster, y = S1, fill=cluster)) +
  geom_bar(stat = "identity") + ggtitle(names(cl.cnt)[1]) +
  annotate("segment", x = 0.5, xend = nrow(DATA)+0.5, y = -0.5, yend = -0.5) +
  annotate("segment", y = 0, yend = 100, x = 0.5, xend = 0.5) +
  theme(plot.title = element_text(size=14, hjust = 0.5),
        panel.background = element_rect(fill="white"),
        legend.position = "none",
        axis.text.x = element_text(size=14),
        axis.title.x = element_blank(), 
        axis.title.y = element_blank(), 
        axis.text.y = element_blank(), 
        axis.ticks.y = element_blank(),
        axis.line.y = element_blank(),
        plot.margin = unit(c(1,-1,1,0), "mm")) +
  scale_y_reverse() + coord_flip(ylim=c(0,round(max(apply(DATA[,-1],2,max)),-1)))

g2 <- ggplot(data = DATA, aes(x = cluster, y = S2, fill=cluster)) +xlab(NULL)+
  geom_bar(stat = "identity") + ggtitle(names(cl.prop)[2]) +
  annotate("segment", x = 0.5, xend = nrow(DATA)+0.5, y = -0.5, yend = -0.5) +
  annotate("segment", y = 0, yend = 100, x = 0.5, xend = 0.5) +
  coord_cartesian(xlim = c(0, round(max(apply(DATA[,-1],2,max))))) +
  theme(plot.title = element_text(size=14, hjust = 0.5),
        panel.background = element_rect(fill="white"),
        legend.position = "none",
        axis.text.x = element_text(size=14),
        axis.title.x = element_blank(), axis.title.y = element_blank(), 
        axis.text.y = element_blank(), axis.ticks.y = element_blank(),
        plot.margin = unit(c(1,0,1,-1), "mm")) +
  coord_flip(ylim=c(0,round(max(apply(DATA[,-1],2,max)),-1)))

library(gridExtra)
gg1 <- ggplot_gtable(ggplot_build(g1))
gg2 <- ggplot_gtable(ggplot_build(g2))
gg.mid <- ggplot_gtable(ggplot_build(g.mid))

grid.arrange(gg1,gg.mid,gg2,ncol=3,widths=c(4/9,1/9,4/9),
             bottom = textGrob("Percentage of cells within-sample (%)",
                               gp=gpar(fontsize=16,font=2)))
```

```{r density_cluster_tab}
write.table(DATA,paste0(OUTDIR,"/cluster_density.tsv"),sep="\t",col.names=TRUE, row.names=FALSE, quote=FALSE)
```

## Additional diagnostics
Whether clustering is driven by mt

```{r umap_mt, dpi=300, fig.width=14, fig.height=6, fig.path="./output/01_integration/", dev=c('png','tiff')}
# Add mitocondrial genes
S[["percent.mt"]] <- PercentageFeatureSet(S, pattern = "^mt-")
d1 <- FeaturePlot(S, reduction = "umap", features = "percent.mt")
d2 <- FeaturePlot(S, reduction = "umap", features = "nCount_RNA")
plot_grid(d1, d2)
```

## Save the Seurat Object
```{r seurat_obj}
saveRDS(S, paste0(OUTDIR,"/S.rds"));
```

## SessionInfo
```{r}
sessionInfo()

{                                                                                                                                                                                                           
sink(file=paste0(OUTDIR,"/sessionInfo.txt"))
print(sessionInfo())
sink()
}
```
