---
title: "Integrated analysis with Seurat of atherosclerotic plaques in the context of Myeloid PHD2 deficiency"
author: "Javier Perales-Patón - javier.perales@bioquant.uni-heidelberg.de - ORCID: 0000-0003-0780-6683" 
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Single-cell transcriptomics profiling of atherosclerotic plaques from a pooled sample of 
Myeloid-lineage PHD2cKO and WT mice were done. Herein we carry out a standard workflow using 
Seurat package, which is used to integrate both samples and identify cell types in the bulk 
sample. Finally, several comparisons with several collections of markers in the biological 
context are performed to test whether the unsupervised clustering recapitulates the expected 
cell populations in the atherosclerotic plaques.

Most of parameters of the analysis used here are based on default settings of Seurat. 
However we have choosen a certain number or principal components based on exploratory analysis on the outcome of those.
We opt to keep low resolution in the unsupervised clustering to identify the supercluster of 
macrophages, later it will be dissected in a second round of unsupervised clustering.

## Setup
We define a random seed number for reproducibility, file structure for the output, 
and what is more important: the parameters of the standard analysis with Seurat. 

### Environment
```{r env}
# Seed number
set.seed(1234)
# Output directory
OUTDIR <- "./01_integration_output/"
if(!dir.exists(OUTDIR)) dir.create(OUTDIR);

# Figures
FIGDIR <- paste0(OUTDIR, "/figures/")
knitr::opts_chunk$set(fig.path=FIGDIR)
knitr::opts_chunk$set(dev=c('png','tiff'))
# Data
DATADIR <- paste0(OUTDIR, "/data/")
if(!dir.exists(DATADIR)) dir.create(DATADIR);
```

## Load libraries
```{r}
suppressPackageStartupMessages(require(Seurat))
suppressPackageStartupMessages(require(GSEABase))
suppressPackageStartupMessages(require(dplyr))
suppressPackageStartupMessages(require(clustree))
suppressPackageStartupMessages(require(cowplot))
suppressPackageStartupMessages(require(genesorteR))
suppressPackageStartupMessages(require(optparse))
# Get some functions for Seurat analysis
source("../src/seurat_fx.R")
source("../src/graphics.R")
```

### Input parameters
```{r}
# Define input parameters
option_list = list(
  make_option(c("--CASE_INPUT"), action="store", 
	      default="../data/sc/CK64/filtered_feature_bc_matrix/", type='character',
              help="cellranger count folder that contains the output matrices"),
  make_option(c("--CASE_SNAME"), action="store", default="PHD2cKO", type='character',
              help="Sample name"),
  make_option(c("--CONTROL_INPUT"), action="store", 
	      default="../data/sc/CK62/filtered_feature_bc_matrix/", type='character',
              help="cellranger count folder that contains the output matrices"),
  make_option(c("--CONTROL_SNAME"), action="store", default="WT", type='character',
              help="Sample name"),
  make_option(c("--NPC"), action="store", default=50, type='numeric',
              help="number of principal components to calculate."),
  make_option(c("--NPC_ANCHOR"), action="store", default=20, type='numeric',
              help="number of principal components to consider for the anchoring."),
  make_option(c("--NPC_CLUSTERING"), action="store", default=25, type='numeric',
              help="Number of principal components to consider for the cell clustering."),
  make_option(c("--RES"), action="store", default=0.1, type='numeric',
              help="Resolution for cell clustering.")
)

# Parse the parameters
opt = parse_args(OptionParser(option_list=option_list))

# Cat the input parameters
cat("[INFO] Input parameters\n", file=stdout())
for(user_input in names(opt)) {
  if(user_input=="help") next;
  cat(paste0("[INFO] ",user_input," => ",opt[[user_input]],"\n"),file = stdout())
  assign(user_input,opt[[user_input]])
}
```

## Load data
Read samples and estimate percentage of mitochondrial gene expression

```{r}
CASE <- getSeuratObject(path = CASE_INPUT, project_name = CASE_SNAME, mt.pattern = "^mt-", min.cells = 5, min.features = 500)
CASE$stim <- CASE_SNAME
CASE <- subset(CASE, subset = nFeature_RNA > 500 & nFeature_RNA < 4000 & percent.mt < 7.5)
CASE <- NormalizeData(CASE, verbose = FALSE)
CASE <- FindVariableFeatures(CASE, selection.method = "vst", nfeatures = 2000)
  
CTRL <- getSeuratObject(path = CONTROL_INPUT, project_name = CONTROL_SNAME, mt.pattern = "^mt-", min.cells = 5, min.features = 500)
CTRL$stim <- CONTROL_SNAME
CTRL <- subset(CTRL, subset = nFeature_RNA > 500 & nFeature_RNA < 4000 & percent.mt < 7.5)
CTRL <- NormalizeData(CTRL, verbose = FALSE)
CTRL <- FindVariableFeatures(CTRL, selection.method = "vst", nfeatures = 2000)
```

## Perform integration
The integration consists on Find anchors between the two samples and perform CCA to integrate both.

```{r}
anchors <- FindIntegrationAnchors(object.list = list(CTRL, CASE), dims = 1:NPC_ANCHOR)
S <- IntegrateData(anchorset = anchors, dims = 1:20)
# Relevel wildtype as ref level
S$stim <- factor(S$stim, levels=c("WT","PHD2cKO"))

# Get some RAM space: we don't need this anymore
rm(CASE, CTRL)
```
We define the Integrated data for cell clustering and assignment
```{r}
DefaultAssay(S) <- "integrated"
```

## Cell clustering and dimensionality reduction
First scale the data and get principal components
```{r}
S <- ScaleData(S, verbose = FALSE)
S <- RunPCA(S, npcs = NPC, verbose = FALSE)
```

Show the elbowplot for the PCs
```{r}
ElbowPlot(S,ndims = NPC) + geom_vline(xintercept = NPC_CLUSTERING, col="red")
```

Perform SNN clustering
```{r}
S <- FindNeighbors(S, reduction = "pca", dims = 1:NPC_CLUSTERING)
S <- FindClusters(S, resolution = seq(from=0.1, to=1.5, by=0.1))
```

```{r clustree, fig.width=14, fig.height=7, dpi=300}
clustree(S, prefix = "integrated_snn_res.")
```

Decide the final resolution: `r RES`
```{r}
S <- FindClusters(S, resolution = RES)
```

Finally, we use UMAP to get 2 reduced dimensions for visualization
```{r}
S <- RunUMAP(S, reduction = "pca", dims = 1:NPC_CLUSTERING)
```

```{r umap_plot, dpi=300}
DimPlot(S, reduction = "umap", group.by = "seurat_clusters", label = TRUE, label.size = 14)
```

```{r umap_condition, dpi=300}
DimPlot(S, reduction = "umap", group.by = "stim",cols = c("grey","red"))
```

### Extract cluster markers
The assay is switched to RNA

```{r}
DefaultAssay(S) <- "RNA"
# cl1.markers <- FindConservedMarkers(S, ident.1 = 1, grouping.var = "stim", verbose = FALSE)
# head(cl1.markers)
# S <- NormalizeData(S, verbose = FALSE) # I am not sure if I have to normalize again
S <- Seurat_scaledata(S)
```

```{r wilcox_test}
up <- setNames(vector("list",length=length(levels(S))), levels(S))
for(idx in names(up)) {
  up.idx <- FindConservedMarkers(S,ident.1 = idx,ident.2 = setdiff(levels(S), idx),
				 grouping.var = "stim",only.pos=T)
  cols_names <- colnames(up.idx)

  # Add two extra cols
  up.idx$cluster <- idx
  up.idx$gene <- rownames(up.idx)

  write.table(up.idx[,c("cluster", "gene", cols_names)],
              file = paste0(DATADIR,"/wilcox_cluster",idx,".tsv"),
              sep="\t",col.names = TRUE, row.names = FALSE, quote=FALSE
  )

  up[[idx]] <- up.idx
}
```

```{r genesorteR}
sg <- sortGenes(S@assays$RNA@data, Idents(S))

#define a small set of markers
mm = getMarkers(sg, quant = 0.975)

#cluster genes and make a heatmap
pp = plotMarkerHeat(sg$inputMat, sg$inputClass, mm$markers, clusterGenes=TRUE, outs = TRUE)

# #the top 25 genes for each cluster by specificity scores
# top_markers = apply(sg$specScore, 2, function(x) names(head(sort(x, decreasing = TRUE), n = 25)))
# write.table(top_markers, paste0(MARKERS.DATADIR,"/top25markers.tsv"),
# 	    sep="\t",row.names = FALSE,col.names = TRUE,quote=FALSE)
# 

# Save the tables of genesorteR stats
write.table(sg$condGeneProb, file=paste0(DATADIR,"/genesorter_condGeneProb.tsv"),sep="\t", row.names=TRUE, col.names = NA, quote=FALSE)
write.table(sg$specScore, file=paste0(DATADIR,"/genesorter_specScore.tsv"),sep="\t", row.names=TRUE, col.names = NA, quote=FALSE)
# Save the object
saveRDS(sg,file = paste0(DATADIR,"/sg.rds"))
```

## Align clustering with previous studies in atherosclerosis and cell type markers
### Internal
```{r heatmap_specscore_top10, fig.width=14, fig.height=9, warning=FALSE}
GSC_markers <- GeneSetCollection(sapply(colnames(sg$specScore),function(idx) GeneSet(names(head(sort(sg$specScore[,idx], decreasing = TRUE), n = 10)),
                                                                                     setName=idx, shortDescription="genesorteR")))

DoHeatmap3(SeuratObject = S, GSC = GSC_markers,
	   fontfamily=fontTXT,
           assay = "RNA", res=NULL, show_hr = FALSE)
```

### External (literature)
```{r heatmap_wirka, fig.width=14,fig.height=14, dpi=300, warning=FALSE}
DoHeatmap3(SeuratObject = S, GSC = getGmt("../data/markers/Wirka_etal2019.gmt"), 
	   fontfamily=fontTXT,
	   assay = "RNA", res=NULL, show_hr=FALSE)
```

```{r heatmap_sluimer, fig.width=14, fig.height=10, dpi=300, warning=FALSE}
DoHeatmap3(SeuratObject = S, GSC = getGmt("../data/markers/markers.gmt"), 
	   fontfamily=fontTXT,
	   assay = "RNA", res=NULL, show_hr=FALSE)
```

```{r heatmap_cochain, fig.width=14, fig.height=14, dpi=300, warning=FALSE}
DoHeatmap3(SeuratObject = S, GSC = getGmt("../data/markers/Cochain_etal2019.gmt"), 
	   fontfamily=fontTXT,
	   assay = "RNA", res=NULL, show_hr=FALSE)
```
```{r heatmap_gu, fig.width=14, fig.height=10, dpi=300, warning=FALSE}
DoHeatmap3(SeuratObject = S, GSC = getGmt("../data/markers/Gu_etal2019.gmt"), 
	   fontfamily=fontTXT,
	   assay = "RNA", res=NULL, show_hr=FALSE)
```

## Density of clusters per conditions

```{r}
cl.cnt<-tapply(S$seurat_clusters,S@meta.data$stim,table)
# sapply(names(cl.cnt), function(idx) sum(S$stim==idx))
cl.prop <- cl.cnt

for(idx in names(cl.prop)) {
  cl.prop[[idx]] <- (cl.prop[[idx]]/sum(S$stim==idx)*100)
}

DATA <- data.frame(cluster=factor(levels(S), levels(S)),
                   S1=as.numeric(cl.prop[[1]]),
                   S2=as.numeric(cl.prop[[2]]))
```

```{r cluster_density, fig.width=7, fig.height=9, dpi=300}
# Source: https://stackoverflow.com/questions/18265941/two-horizontal-bar-charts-with-shared-axis-in-ggplot2-similar-to-population-pyr
library(grid)
g.mid<-ggplot(DATA,aes(x=1,y=cluster))+geom_text(aes(label=as.character(DATA$cluster)), size=14)+
  geom_segment(aes(x=0.94,xend=0.96,yend=cluster))+
  geom_segment(aes(x=1.04,xend=1.065,yend=cluster))+
  ggtitle("Cluster")+
  ylab(NULL)+
  scale_x_continuous(expand=c(0,0),limits=c(0.94,1.065))+
  theme(plot.title = element_text(size=14),
        panel.background = element_rect(fill="white"),
        axis.title=element_blank(),
        panel.grid=element_blank(),
        axis.text.y=element_blank(),
        axis.ticks.y=element_blank(),
        axis.text.x=element_text(color=NA, size=14),
        axis.ticks.x=element_line(color=NA),
        plot.margin = unit(c(1,-1,1,-1), "mm"))

g1 <- ggplot(data = DATA, aes(x = cluster, y = S1, fill=cluster)) +
  geom_bar(stat = "identity") + ggtitle(names(cl.cnt)[1]) +
  annotate("segment", x = 0.5, xend = nrow(DATA)+0.5, y = -0.5, yend = -0.5) +
  annotate("segment", y = 0, yend = 100, x = 0.5, xend = 0.5) +
  theme(plot.title = element_text(size=14, hjust = 0.5),
        panel.background = element_rect(fill="white"),
        legend.position = "none",
        axis.text.x = element_text(size=14),
        axis.title.x = element_blank(), 
        axis.title.y = element_blank(), 
        axis.text.y = element_blank(), 
        axis.ticks.y = element_blank(),
        axis.line.y = element_blank(),
        plot.margin = unit(c(1,-1,1,0), "mm")) +
  scale_y_reverse() + coord_flip(ylim=c(0,round(max(apply(DATA[,-1],2,max)),-1)))

g2 <- ggplot(data = DATA, aes(x = cluster, y = S2, fill=cluster)) +xlab(NULL)+
  geom_bar(stat = "identity") + ggtitle(names(cl.prop)[2]) +
  annotate("segment", x = 0.5, xend = nrow(DATA)+0.5, y = -0.5, yend = -0.5) +
  annotate("segment", y = 0, yend = 100, x = 0.5, xend = 0.5) +
  coord_cartesian(xlim = c(0, round(max(apply(DATA[,-1],2,max))))) +
  theme(plot.title = element_text(size=14, hjust = 0.5),
        panel.background = element_rect(fill="white"),
        legend.position = "none",
        axis.text.x = element_text(size=14),
        axis.title.x = element_blank(), axis.title.y = element_blank(), 
        axis.text.y = element_blank(), axis.ticks.y = element_blank(),
        plot.margin = unit(c(1,0,1,-1), "mm")) +
  coord_flip(ylim=c(0,round(max(apply(DATA[,-1],2,max)),-1)))

library(gridExtra)
gg1 <- ggplot_gtable(ggplot_build(g1))
gg2 <- ggplot_gtable(ggplot_build(g2))
gg.mid <- ggplot_gtable(ggplot_build(g.mid))

grid.arrange(gg1,gg.mid,gg2,ncol=3,widths=c(4/9,1/9,4/9),
             bottom = textGrob("Percentage of cells within-sample (%)",
                               gp=gpar(fontsize=16,font=2)))
```

```{r density_cluster_tab}
write.table(DATA,paste0(DATADIR,"/cluster_density.tsv"),sep="\t",col.names=TRUE, row.names=FALSE, quote=FALSE)
```

## Additional diagnostics
Whether clustering is driven by mt

```{r umap_mt, dpi=300, fig.width=14, fig.height=6}
# Add mitocondrial genes
S[["percent.mt"]] <- PercentageFeatureSet(S, pattern = "^mt-")
d1 <- FeaturePlot(S, reduction = "umap", features = "percent.mt")
d2 <- FeaturePlot(S, reduction = "umap", features = "nCount_RNA")
plot_grid(d1, d2)
```

## Save the Seurat Object
```{r seurat_obj}
saveRDS(S, paste0(DATADIR,"/S.rds"));
```

## SessionInfo
```{r}
sessionInfo()

{                                                                                                                                                                                                           
sink(file=paste0(OUTDIR,"/sessionInfo.txt"))
print(sessionInfo())
sink()
}
```
