---
title: "Investigation of the role of PHD2 in-vivo atherogenesis in Macrophages"
author: "Javier Perales-Patón - javier.perales@bioquant.uni-heidelberg.de"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load libraries
```{r}
suppressPackageStartupMessages(require(Seurat))
suppressPackageStartupMessages(require(GSEABase))
suppressPackageStartupMessages(require(cowplot))
suppressPackageStartupMessages(require(genesorteR))
suppressPackageStartupMessages(require(fgsea))
suppressPackageStartupMessages(require(ggplot2))
suppressPackageStartupMessages(require(ggrepel))
suppressPackageStartupMessages(require(AUCell))
suppressPackageStartupMessages(require(viper))
suppressPackageStartupMessages(require(purrr))
suppressPackageStartupMessages(require(dplyr))
source("../src/graphics.R")
source("../src/seurat_fx.R")
```
### Setting-up environment
The environment will be set with a random seed number for reproducibility and an output folder
for processed data and figures.

```{r env}
# Seed number
set.seed(1234)
# Output directory
OUTDIR <- "./output/03_macrophages"
if(!dir.exists(OUTDIR)) dir.create(OUTDIR);
# Input data
sobj <- "./output/02_assignment/S.rds"
```


## Load data
Read the Seurat Object from second step.
```{r load_seurat}
if(file.exists(sobj)) {
	S <- readRDS(sobj)
} else {
	stop("ERROR: Seurat object does not exist. Run 02.rmd to generate it.")
}
```

We also load the ortholog information for functional analysis
```{r}
# Lib of hsa2mmu orthologs
mmu2hsa <- readRDS("../data/Gene_annotation/hgnc2mgi_hsa2mmu.rds")
```

## Comparison between PHD2 functional perturbation
The in-vivo model carries a PHD2 knock-out in the myeloid lineage with an effectivity of 30% of the cells.
The section will investigate if the penetrance of this gene perturbation is enough to be capture in the 
average population of the mouse replicate.

### Differential gene expression
```{r scatterplot_macrophages_PHD2cKO_vs_PHD2wt, fig.width=5, fig.height=5, dpi=300, fig.path="./output/03_macrophages/", dev=c('png','tiff')}
MARKERS.OUTDIR <- paste0(OUTDIR,"/Macrophages_avg_PHD2cKO_vs_PHD2wt")
if(!dir.exists(MARKERS.OUTDIR)) dir.create(MARKERS.OUTDIR);

S1 <- S[, S@active.ident=="Macrophage"]

Idents(S1) <- "stim"
avg <- log1p(AverageExpression(S1, verbose = FALSE)$RNA)
  
rnk <- setNames(avg[,"PHD2cKO"] - avg[,"WT"], rownames(avg))
top <- names(sort(abs(rnk),decreasing = TRUE)[1:15])
  
ggplot(avg, aes(WT, PHD2cKO)) + geom_point(alpha=0.6) + ggtitle(label = "Macrophages") + 
	xlab("WT") + ylab("PHD2cKO")+
	geom_label_repel(data=avg[top,], aes(label=top), size=5) + 
	theme_cowplot() +
      theme(axis.title = element_text(size=22),
            title = element_text(size=22, hjust = 0.5),
            axis.text = element_text(size=18),
            axis.title.x = element_text(size=18),
            axis.title.y = element_text(size=18))
    
  
dge <- FindMarkers(object = S1, ident.1 = "PHD2cKO", ident.2 = "WT", 
		  logfc.threshold=0, min.pct=0)
cols_names <- colnames(dge)
# Add two extra cols
dge$cluster <- "Macrophage PHD2cKO vs PHD2wt"
dge$gene <- rownames(dge)

dge <- dge[,c("cluster", "gene", cols_names)]
   
write.table(dge,
            file = paste0(MARKERS.OUTDIR,"/Macrophages_avg_PHD2cKO_vs_PHD2wt_DEGs.tsv"),
            sep="\t",col.names = TRUE, row.names = FALSE, quote=FALSE
)
```

There are these number of differentially expressed genes between the two conditions:
```{r}
table(sign(dge$avg_logFC) * as.integer(dge$p_val_adj < 0.05))
```

We could conclude that there are not large differences in the average gene expression of 
macrophages based on condition (PHD2 KO). Next efforts will be made to recover transcriptional 
programmes on this regulation and try to dissect cell populations

### Functional analysis
We create a ranking of differentially expressed genes for the enrichment analysis.
```{r}
rnk <- setNames(sign(dge$avg_logFC) * -log10(dge$p_val_adj), dge$gene)
```

```{r}
# GO analysis
GO <- getGmt("../data/MSigDB/c5.bp.v7.0.symbols.gmt")
GO <- lapply(geneIds(GO), function(z) unlist(sapply(z, function(j) mmu2hsa[[j]])))

GO.res <- fgsea(pathways = GO, stats = rnk, nperm=10000) 
GO.res <- GO.res[order(GO.res$padj, decreasing=FALSE),]
head(GO.res)
```

```{r}
# HALLMARK
H <- getGmt("../data/MSigDB/h.all.v6.2.symbols.gmt")
H <- lapply(geneIds(H), function(z) unlist(sapply(z, function(j) mmu2hsa[[j]])))

H.res <- fgsea(pathways = H, stats = rnk, nperm=10000) 
H.res <- H.res[order(H.res$padj, decreasing=FALSE),]
head(H.res)
```

It seems that hypoxia is up-regulated in PHD2-cKO condition, but the signal is 
not strong enough. Given the biology of the two conditions, it could be the case 
that the PHD2-wt is also responding to hypoxia because of the atherogenesis, enmasking 
the output of the contrast. Next section will study this.

## Dissection of hypoxia response and PHD2 functional impairment
Herein a transcriptional dissection of hypoxia is performed driven by the
hypothesis that PHD2-KO perturbation triggers a strong hypoxia response. It is expected
that the PHD2-cKO condition presents an over-representation of cells responding to hypoxia.

### Dorothea focused on Hif1a
Text

```{r vln_macrophages_dorothea_hif1a, fig.width=6, fig.height=7, dpi=300, fig.path="./output/03_macrophages/", dev=c('png','tiff')}
# load regulons
df2regulon <- function(df, regulator_name="tf") {
  regulon = df %>% split(.[regulator_name]) %>% map(function(dat) {
    targets = setNames(dat$mor, dat$target)
    likelihood = dat$likelihood
    list(tfmode = targets, likelihood = likelihood)
  })
  return(regulon)
}

regulon.df <- read.table("../data/Prior/dorothea_regulon_mouse_v1.csv", sep=",", header=TRUE, stringsAsFactors = FALSE)
regul <- df2regulon(df=regulon.df)

# Calculate TF activities
TF <- viper(eset = as.matrix(S@assays$RNA@data), regulon = regul,
              nes = T, method = "none", minsize = 4,
              eset.filter = F, adaptive.size = F)
  
# Add them as metadata
stopifnot(colnames(S) == colnames(TF))
S$Hif1a_activity <- TF["Hif1a",]
rm(TF)

# Visualization focused on Macrophages population
  #VlnPlot(S[, Idents(S)=="Macrophage"],features = "Hif1a_activity", group.by = "stim")
VlnPlot.stim(S[,Idents(S)=="Macrophage"], 
     meta.feature = "Hif1a_activity", ylab="Hif1a activity", 
     fontTXT=fontTXT)

```

### PROGENy focused on hypoxia response
Text

```{r vln_progeny_hypoxia_macrophages, fig.width=6, fig.height=7, dpi=300, fig.path="./output/03_macrophages/", dev=c('png','tiff')}
### Progeny ####
progeny.mat <- read.table("../data/Prior/progeny_matrix_mouse_v1.txt",sep=",",header=TRUE)
rownames(progeny.mat) <- progeny.mat$X
progeny.mat <- progeny.mat[which(colnames(progeny.mat)!="X")]
progeny.mat <- as.matrix(progeny.mat)

common <- intersect(rownames(S), rownames(progeny.mat))
  
prog <- t(as.matrix(S@assays$RNA@data[common,])) %*% progeny.mat[common,]
rn <- rownames(prog)
prog <- apply(prog,2,scale)
rownames(prog) <- rn
prog <- t(prog)
  
stopifnot(colnames(S) == colnames(prog))
S$Hypoxia_response <- prog["Hypoxia",]
rm(common,prog)

# Visualization focused on Macrophages population
 # VlnPlot(S[, Idents(S)=="Macrophage"],features = "Hypoxia_response", group.by = "stim")
VlnPlot.stim(S[,Idents(S)=="Macrophage"], 
     meta.feature = "Hypoxia_response", ylab="Hypoxia response (Score)", 
     fontTXT=fontTXT)


```

### Enrichment of in-vitro PHD2-KO signature
```{r define_signature}
# Load in-vitro signature from the bulk analysis
invitro.dge <- read.table("../Analysis_invitro_bulk/output/01_bulk_dge/MC_PHD2_diffexpr.tsv", 
			  sep="\t", header=TRUE, stringsAsFactors = FALSE)
invitro.sign <- sort(setNames(invitro.dge$t,invitro.dge$genes),decreasing = TRUE)
# Get common genes
common <- intersect(rownames(S), names(invitro.sign))
# Discard certain genes that are driving
avoid_genes <- c("Bnip3", "Spp1")
common <- setdiff(common, avoid_genes)
# Define signature and show
invitro.gs <- names(sort(invitro.sign[common],decreasing = TRUE))[1:50]
print(invitro.gs)
```

```{r vln_aucell_phd2ko_macrophages, fig.width=6, fig.height=7, dpi=300, fig.path="./output/03_macrophages/", dev=c('png','tiff')}
# Build rankings for AUC
cells_rankings <- AUCell_buildRankings(as.matrix(S@assays$RNA@data))
# Use top N (e.g. 50) to calculate enrichment at single-cell level
cells_AUC <- AUCell_calcAUC(list("invitro"=invitro.gs), cells_rankings, aucMaxRank=nrow(cells_rankings)*0.05)
# Explore thresholds
cells_assignment <- AUCell_exploreThresholds(cells_AUC, plotHist=FALSE, assign=TRUE) 
# Get AUCell scores  
AUC.sign <- getAUC(cells_AUC)[1,]
stopifnot(colnames(S) == names(AUC.sign))
S$PHD2cKO <- AUC.sign
rm(AUC.sign, cells_rankings, cells_AUC, cells_assignment)

# Visualization focused on Macrophages population
 # VlnPlot(S[, Idents(S)=="Macrophage"],features = "PHD2cKO", group.by = "stim")
VlnPlot.stim(S[,Idents(S)=="Macrophage"], 
     meta.feature = "PHD2cKO", ylab="Expression of PHD2cKO signature (AUC)", 
     fontTXT=fontTXT)
```

### Pair-wise comparison of these markers of hypoxia response

```{r pairs_hypoxia_macrophages, fig.width=9, fig.height=9, dpi=300, fig.path="./output/03_macrophages/", dev=c('png','tiff')}
# Define colors for ggplot
cond.cols <- c("WT"="grey", "PHD2cKO"="red")

panel.cor <- function(x, y, digits=2, prefix="", cex.cor, family=fontTXT, ...)
{
  usr <- par("usr"); on.exit(par(usr))
  par(usr = c(0, 1, 0, 1))
  r <- abs(cor(x, y, method="pearson"))
  txt <- format(c(r, 0.123456789), digits=digits)[1]
  txt <- paste(prefix, txt, sep="")
  if(missing(cex.cor)) cex.cor <- 0.8/strwidth(txt)
  text(0.5, 0.5, txt, cex = cex.cor * r, family=family)
}

# Visualization
pairs(S@meta.data[Idents(S)=="Macrophage",
      		c("Hif1a_activity","Hypoxia_response","PHD2cKO")],
      lower.panel=panel.smooth, upper.panel=panel.cor, cex.labels = 2.6, cex.axis=2.3, las=1,
      family=fontTXT,
      pch=21,
      bg=cond.cols[S@meta.data[Idents(S)=="Macrophage", "stim"]])
```

We could conclude that PHD2-cKO condition presents a skewed distribution of cells towards 
hypoxia response as compared to PHD2-wt condition.

### Stratify

```{r}
cutoff <- quantile(S@meta.data[Idents(S)=="Macrophage" & S$stim=="PHD2cKO", "PHD2cKO"], 
		   probs=0.75) 

S$PHD2cKO_class <- NA

S$PHD2cKO_class[Idents(S)=="Macrophage"] <- ifelse(S$PHD2cKO[Idents(S)=="Macrophage"] > cutoff,
						   "High", "Low")

S$PHD2cKO_class2 <- paste0(S$stim,"_",S$PHD2cKO_class)

```

```{r vln_aucell_phd2ko_macrophages_wCutoff, fig.width=6, fig.height=7, dpi=300, fig.path="./output/03_macrophages/", dev=c('png','tiff')}
# Visualization focused on Macrophages population
VlnPlot.stim(S[,Idents(S)=="Macrophage"], 
     meta.feature = "PHD2cKO", ylab="Expression of PHD2cKO signature (AUC)", 
     fontTXT=fontTXT) + geom_hline(yintercept=cutoff)
```



```{r}
S1 <- S[, Idents(S)=="Macrophage"] 
Idents(S1) <- "PHD2cKO_class2"
dge <- FindMarkers(S1, ident.1 = "PHD2cKO_High", ident.2 = "WT_Low",
		   min.pct = 0, logfc.threshold = 0)
```


```{r volcano_PHD2cKOsign_high_vs_low, dpi=300, fig.width=8, fig.height=6, fig.path='./output/03_macrophages/', dev = c('png','tiff')}
Mac_interesting_genes <- getGmt("../data/Prior/MC_PHD2_diff.gmt")
Mac_interesting_genes <- unlist(geneIds(Mac_interesting_genes))
if(any(!Mac_interesting_genes %in% rownames(S1))) {
	cat(paste("The following genes are not present in the gene expr matrix:","\n",
		  paste(setdiff(Mac_interesting_genes, rownames(S1)), collapse=","),
		  "\n"),
		  file=stdout())
	Mac_interesting_genes <- intersect(Mac_interesting_genes, rownames(S1))
}

dge$significant <- ifelse(dge$p_val_adj < 0.05,"Adj.P-value<0.05","n.s.")
dge$genes <- rownames(dge)
dge$show <- dge$genes %in% Mac_interesting_genes
 
p = ggplot(dge, aes(avg_logFC, -log10(p_val+0.001))) +
    geom_point(aes(col=significant), alpha=0.6) +
    scale_color_manual(values=c("red", "black"))
  
  
p+geom_label_repel(data=dge[dge$show,],
                          aes(label=genes), family=fontTXT, size=8,
                          force=2,
			  xlim = c(2.5,3.0),
                          nudge_y=0.05, direction = "y", segment.size = 0.5) + 
    coord_cartesian(xlim = c(-1.3, 2.3), clip = "off") +
    theme_cowplot() +
    xlab("avg log2FC") + ylab("-log10(pvalue + 0.001)") +
    theme(text = element_text(family=fontTXT, size=20),
	  legend.text = element_text(family=fontTXT, size=20),
	  axis.text = element_text(family=fontTXT, size=20),
    	  plot.margin = unit(c(0.3, 2.6, 0.1, 0.1), "cm"),
	  legend.position = "bottom",
	  legend.direction = "horizontal"
	  )
```

## Save the Seurat Object
```{r seurat_obj}
saveRDS(S, paste0(OUTDIR,"/S.rds"));
```

## SessionInfo
```{r}
sessionInfo()

{                                                                                                                                                                                                           
sink(file=paste0(OUTDIR,"/sessionInfo.txt"))
print(sessionInfo())
sink()
}
```
