---
title: "Analysis invivo PFKFB3 - plaque cells from LDLrlysMCRE mice model"
author: "Javier Perales-Patón - javier.perales@bioquant.uni-heidelberg.de - ORCID: 0000-0003-0780-6683" 
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
## Setup
We define a random seed number for reproducibility, file structure for the output, 
and load essential libraries

### Environment
```{r env}
# Seed number
set.seed(1234)
# Output directory
OUTDIR <- "./LDLrlysMCRE-mice/"
if(!dir.exists(OUTDIR)) dir.create(OUTDIR);

# Figures
FIGDIR <- paste0(OUTDIR, "/figures/")
knitr::opts_chunk$set(fig.path=FIGDIR)
knitr::opts_chunk$set(dev=c('png','tiff'))
knitr::opts_chunk$set(dpi=300)
# Data
DATADIR <- paste0(OUTDIR, "/data/")
if(!dir.exists(DATADIR)) dir.create(DATADIR);
```


### Load libraries
```{r}
suppressPackageStartupMessages(require(Seurat))
suppressPackageStartupMessages(require(cowplot))
suppressPackageStartupMessages(require(ggplot2))
suppressPackageStartupMessages(require(purrr))
suppressPackageStartupMessages(require(dplyr))
suppressPackageStartupMessages(require(viper))
suppressPackageStartupMessages(require(rcompanion))
source("../src/graphics.R")
source("../src/seurat_fx.R")
source("../src/wilcox_fx.R")
```

## Load data
Read the Seurat Object from the in-vivo analysis. We use the same annotation
```{r load_seurat}
readSeuratObject <- function(sobj) {
	# Read 
	if(file.exists(sobj)) {
		S <- readRDS(sobj)
	} else {
		stop("ERROR: Seurat object does not exist.")
	}
	return(S)
}
```

Read the SeuratObject with all plaque cells
```{r}
# Input data : from prev
S <- readSeuratObject("../LDLrlysMCRE_clustering/02_identity_output/data/S.rds")
S$Annotation.Level.1 <- Idents(S)
```

Read the subset of myeloid cells which were re-cluster in another iteration, annotated afterwards.
```{r}
M <- readSeuratObject("../LDLrlysMCRE_clustering/03_myeloidlineage_output/data/M.rds")
M$Annotation.Level.2<- Idents(M)
```


## Subsetting data
Since the latter is contained in the former, we use the same criteria for dropping out
the cells from the latter in the former: Discard Lyz2+ from whole tissue.
```{r}
myeloid <- c("Megakaryocyte", "Thrombocyte", "Erythrocyte", "Mast cell",
	     "Basophil", "Neutrophil", "Eosinophil", "Monocyte", "Macrophage")
grep(paste(myeloid, collapse="|"), levels(S), value=TRUE)

S <- S[, !Idents(S) %in% c("Mixed_Neutrophil-DC-Monocytes", "Macrophage", "n.a.")]
```

Finally we have two subsets of the cells:
- Lyz2- cells: including fibroblasts, SMC, EC, T-cell and B-cell

```{r show_Lyz2pos}
# Classes
levels(S)
# Cell numbers
table(S$Annotation.Level.1)
```

- Lyz2+ cells: Myeloid leukocytes.
```{r show_Lyz2neg}
# Classes
levels(M)
# Cell numbers
table(M$Annotation.Level.2)
```

Note, doing so two clusters of cells are dropping from the analysis:
- A small cluster with double marker expression of myeloid cells called "n.a." (not assigned).
  Potentially those are doublets, but I couldn't identify them with regular methods because
  they are very poor represented.
- Myeloid proliferating cells (mixed of many cell types).


In addition, we subset the data by condition. Actually the data contains two conditions,
WT (control) mice and myeloid-PHD2cKO mice. We are mainly interested in the control since
that PHD2cKO perturbation was profiled for other study. Thus we are reusing the control here.

```{r}
WT <- merge(subset(S, stim=="WT"), subset(M, stim=="WT"))
WT$is_myeloid <- ifelse(is.na(WT$Annotation.Level.2), FALSE, TRUE)
```

```{r reorder}
# Sort them by, 1st Macrophages, then the rest, finally the odd ones
last_ids<- c("EC1", "EC2", "Fibroblast", "SMC")
ids <- c(sort(grep("-Mac", levels(WT), value=TRUE)),
	 setdiff(sort(grep("-Mac", levels(WT), value=TRUE, invert=TRUE)),
		 last_ids),
	 last_ids)
Idents(WT) <- factor(as.character(Idents(WT)),
		      levels=ids)
# We show the total number of cells
ncol(WT)
# And the sample size of each group
table(Idents(WT))
```

```{r LDLrlysMCRE_WT_Lyz2_ViolinPlots, fig.width=6, fig.height=5, warning=FALSE}
DefaultAssay(WT) <- "RNA"
plotVln(SeuratObject = WT, gene="Lyz2", meta=NULL,
 	stats=NULL,
	vlnsplit = FALSE, fontTXT,  nCell.y=-0.3, pt.alpha=0.7) +
        NoLegend()
```

We repeat same plot but w/o dots (observations)
```{r LDLrlysMCRE_WT_Lyz2_ViolinPlots_woDots, fig.width=6, fig.height=5, warning=FALSE}
DefaultAssay(WT) <- "RNA"
plotVln(SeuratObject = WT, gene="Lyz2", meta=NULL,
 	stats=NULL,
	vlnsplit = FALSE, fontTXT,  nCell.y=-0.3, pt.alpha=0) +
        NoLegend()
```

## UMAP
We have subset the data so we do a selection of variables, scaling etc for umap
```{r}
DefaultAssay(WT) <- "RNA"
# WT <- NormalizeData(WT, verbose=FALSE)
WT <- ScaleData(WT, verbose=FALSE)
WT <- FindVariableFeatures(WT, 
			   selection.method = "vst",
			   nfeatures = 2000, verbose = FALSE)
WT <- RunPCA(WT, npcs = 30, verbose = FALSE)
# ElbowPlot(M.i)
WT <- RunUMAP(WT, reduction = "pca", dims = 1:15)
```

```{r LDLrlysMCRE-mice_WT_UMAP, fig.width=7, fig.height=5}
p1 <- DimPlot(WT, label=TRUE, repel = TRUE)
print(p1)
```
## Dissection of hypoxia response

### Dorothea focused on Hif1a
We calculate Hif1a transcription factor activities using dorothea. 

```{r dorothea_hif1a, fig.width=6, fig.height=7, dpi=300, warning=FALSE}
# load regulons
df2regulon <- function(df, regulator_name="tf") {
  regulon = df %>% split(.[regulator_name]) %>% map(function(dat) {
    targets = setNames(dat$mor, dat$target)
    likelihood = dat$likelihood
    list(tfmode = targets, likelihood = likelihood)
  })
  return(regulon)
}

regulon.df <- read.table("../data/Prior/dorothea_regulon_mouse_v1.csv", sep=",", 
			 header=TRUE, stringsAsFactors = FALSE)
regul <- df2regulon(df=regulon.df)

# Calculate TF activities
TF <- viper(eset = as.matrix(WT@assays$RNA@data), regulon = regul,
              nes = T, minsize = 4,
              eset.filter = F, adaptive.size = F,
	      verbose=FALSE)
  
# Add them as metadata
stopifnot(colnames(WT) == colnames(TF))
WT$Hif1a_activity <- TF["Hif1a",]
rm(TF)
```

```{r Hif1a_strata}
WT$Hif1a_strata <- NA
for(cell in levels(WT)) {
	idx <- WhichCells(WT, idents=cell)
	Q3 <- quantile(WT$Hif1a_activity[idx], probs=0.75)
	WT$Hif1a_strata[idx] <- ifelse(WT$Hif1a_activity[idx] > Q3,
					 "High_Hif1a", "Low_Hif1a")
}
WT$Hif1a_strata <- factor(WT$Hif1a_strata, levels=c("Low_Hif1a", "High_Hif1a"))
```

### PROGENy focused on hypoxia response
We calculate a score of hypoxia response using progeny as well.

```{r vln_progeny_hypoxia, fig.width=6, fig.height=7, dpi=300, warning=FALSE}
### Progeny ####
progeny.mat <- read.table("../data/Prior/progeny_matrix_mouse_v1.txt",sep=",",header=TRUE)
rownames(progeny.mat) <- progeny.mat$X
progeny.mat <- progeny.mat[which(colnames(progeny.mat)!="X")]
progeny.mat <- as.matrix(progeny.mat)

common <- intersect(rownames(WT), rownames(progeny.mat))
  
prog <- t(as.matrix(WT@assays$RNA@data[common,])) %*% progeny.mat[common,]
rn <- rownames(prog)
prog <- apply(prog,2,scale)
rownames(prog) <- rn
prog <- t(prog)
  
stopifnot(colnames(WT) == colnames(prog))
WT$Hypoxia_response <- prog["Hypoxia",]
rm(common,prog)
```

```{r hypoxia_withinpop_scaled}
WT$Hypoxia_scaled <- NA
for(cell in levels(WT)) {
	idx <- WhichCells(WT, idents=cell)
	WT$Hypoxia_scaled[idx] <- scale(WT$Hypoxia_response[idx])
}
```

```{r hypoxia_strata}
WT$Hypoxia_strata <- NA
for(cell in levels(WT)) {
	idx <- WhichCells(WT, idents=cell)
	Q3 <- quantile(WT$Hypoxia_response[idx], probs=0.75)
	WT$Hypoxia_strata[idx] <- ifelse(WT$Hypoxia_response[idx] > Q3,
					 "High_hypoxia", "Low_hypoxia")
}
WT$Hypoxia_strata<- factor(WT$Hypoxia_strata, levels=c("Low_hypoxia", "High_hypoxia"))
```

## Results

```{r gene}
gene <- "Pfkfb3"
genes <- paste0("Pfkfb",1:4) 
```

### Pfkfb3
```{r LDLrlysMCRE-mice_WT_PFKFB3_FeaturePlots, fig.width=6, fig.height=5}
p2 <- FeaturePlot(WT, features=gene, pt.size = 0.3)
print(p2)
```

```{r LDLrlysMCRE-mice_WT_PFKFB3_UMAPFeaturePlots, fig.width=14, fig.height=5}
plot_grid(p1,p2, rel_widths = c(0.55,0.45))
```

```{r LDLrlysMCRE-mice_WT_PFKFB3_DotPlots, fig.width=5, fig.height=5}
DotPlot(WT, features=gene) + 
	xlab("") + ylab("Cell type")
```

```{r LDLrlysMCRE-mice_WT_PFKFB3_ViolinPlots, fig.width=6, fig.height=5, warning=FALSE}
# VlnPlot(WT, features=gene, pt.size=0.7) + NoLegend()
plotVln(SeuratObject = WT, gene="Pfkfb3", meta=NULL,
	stats=NULL,
	vlnsplit = FALSE, fontTXT,  nCell.y=-0.1, pt.alpha=0.8) +
        NoLegend()
```

### Pfkfb isoforms
```{r LDLrlysMCRE-mice_WT_PFKFBiso_FeaturePlots, fig.width=7.5, fig.height=6}
p2 <- FeaturePlot(WT, features=genes, pt.size = 0.3)
print(p2)
```

```{r LDLrlysMCRE-mice_WT_PFKFBiso_UMAPFeaturePlots, fig.width=14, fig.height=5}
plot_grid(p1,p2, rel_widths = c(0.53,0.47))
```

```{r LDLrlysMCRE-mice_WT_PFKFBiso_DotPlots}
DotPlot(WT, features=genes) + 
	xlab("Pfkfb isoforms") + ylab("Cell type")
```

```{r LDLrlysMCRE-mice_WT_PFKFBiso_ViolinPlots, fig.width=10, fig.height=10, warning=FALSE}
# VlnPlot(WT, features=genes, ncol=length(genes)/2, pt.size=0.7)
plot_grid(plotlist=sapply(genes,function(z) {
		plotVln(SeuratObject = WT, gene=z, meta=NULL,
		  stats=NULL,
		  vlnsplit = FALSE, fontTXT,  nCell.y=-0.1, pt.alpha=0.8) +
			  theme(axis.text.x=element_text(size=16)) + 
			  NoLegend()
		 }, simplify=FALSE), ncol = length(genes)/2)
```


### Pfkfb3 High Hypoxia/Hif1a

#### Hypoxia
```{r LDLrlysMCRE-mice_WThypoxiaHighvsLow_PFKFB3_stats}
WT$stim <- WT$Hypoxia_strata
wPfkfb3_stats <- t(sapply(levels(WT), function(cell) {
				   cellIds <- WhichCells(WT, idents=cell)
				   # The test
				   wilcox.test_stats(xy=WT@assays$RNA@data["Pfkfb3", cellIds],
						     gr=WT$stim[cellIds])
			   }))
wPfkfb3_stats <- as.data.frame(wPfkfb3_stats)
wPfkfb3_stats$adjpval <- p.adjust(wPfkfb3_stats$pvalue, method="fdr")
wPfkfb3_stats$significance <- tagSignif(wPfkfb3_stats$adjpval)
print(wPfkfb3_stats)

RES <- wPfkfb3_stats[, c("r", "significance")]
RES$r <- format(round(RES$r, 2), nsmall=2)
colnames(RES)[ncol(RES)] <- ""
```


```{r LDLrlysMCRE-mice_WThypoxiaHighvsLow_PFKFB3, fig.width=10, fig.height=7, warning=FALSE}
plotVln(SeuratObject = WT, gene="Pfkfb3", meta=NULL,
	stats=RES,
	vlnsplit = TRUE, fontTXT,  nCell.y=-0.1, pt.alpha=0.4) +
      guides(fill= guide_legend(nrow=1, byrow=TRUE)) 

```

#### Hif1a

```{r LDLrlysMCRE-mice_WTHif1aHighvsLow_PFKFB3_stats}
WT$stim <- WT$Hif1a_strata
wPfkfb3_stats <- t(sapply(levels(WT), function(cell) {
				   cellIds <- WhichCells(WT, idents=cell)
				   # The test
				   wilcox.test_stats(xy=WT@assays$RNA@data["Pfkfb3", cellIds],
						     gr=WT$stim[cellIds])
			   }))
wPfkfb3_stats <- as.data.frame(wPfkfb3_stats)
wPfkfb3_stats$adjpval <- p.adjust(wPfkfb3_stats$pvalue, method="fdr")
wPfkfb3_stats$significance <- tagSignif(wPfkfb3_stats$adjpval)
print(wPfkfb3_stats)

RES <- wPfkfb3_stats[, c("r", "significance")]
RES$r <- format(round(RES$r, 2), nsmall=2)
colnames(RES)[ncol(RES)] <- ""
```

```{r LDLrlysMCRE-mice_WTHif1aHighvsLow_PFKFB3, fig.width=10, fig.height=7, warning=FALSE}
plotVln(SeuratObject = WT, gene="Pfkfb3", meta=NULL,
	stats=RES,
	vlnsplit = TRUE, fontTXT,  nCell.y=-0.1, pt.alpha=0.4) +
      guides(fill= guide_legend(nrow=1, byrow=TRUE))

```

### Pfkfb4 High Hypoxia/Hif1a


#### Hypoxia
```{r LDLrlysMCRE-mice_WThypoxiaHighvsLow_PFKFB4_stats}
WT$stim <- WT$Hypoxia_strata
wPfkfb4_stats <- t(sapply(levels(WT), function(cell) {
				   cellIds <- WhichCells(WT, idents=cell)
				   # The test
				   wilcox.test_stats(xy=WT@assays$RNA@data["Pfkfb4", cellIds],
						     gr=WT$stim[cellIds])
			   }))
wPfkfb4_stats <- as.data.frame(wPfkfb4_stats)
wPfkfb4_stats$adjpval <- p.adjust(wPfkfb4_stats$pvalue, method="fdr")
wPfkfb4_stats$significance <- tagSignif(wPfkfb4_stats$adjpval)
print(wPfkfb4_stats)

RES <- wPfkfb4_stats[, c("r", "significance")]
RES$r <- format(round(RES$r, 2), nsmall=2)
colnames(RES)[ncol(RES)] <- ""
```


```{r LDLrlysMCRE-mice_WThypoxiaHighvsLow_PFKFB4, fig.width=10, fig.height=7, warning=FALSE}
plotVln(SeuratObject = WT, gene="Pfkfb4", meta=NULL,
	stats=RES,
	vlnsplit = TRUE, fontTXT,  nCell.y=-0.1, pt.alpha=0.4) +
      guides(fill= guide_legend(nrow=1, byrow=TRUE)) 

```

#### Hif1a
```{r LDLrlysMCRE-mice_WTHif1aHighvsLow_PFKFB4_stats}
WT$stim <- WT$Hif1a_strata
wPfkfb4_stats <- t(sapply(levels(WT), function(cell) {
				   cellIds <- WhichCells(WT, idents=cell)
				   # The test
				   wilcox.test_stats(xy=WT@assays$RNA@data["Pfkfb4", cellIds],
						     gr=WT$stim[cellIds])
			   }))
wPfkfb4_stats <- as.data.frame(wPfkfb4_stats)
wPfkfb4_stats$adjpval <- p.adjust(wPfkfb4_stats$pvalue, method="fdr")
wPfkfb4_stats$significance <- tagSignif(wPfkfb4_stats$adjpval)
print(wPfkfb4_stats)

RES <- wPfkfb4_stats[, c("r", "significance")]
RES$r <- format(round(RES$r, 2), nsmall=2)
colnames(RES)[ncol(RES)] <- ""
```


```{r LDLrlysMCRE-mice_WTHif1aHighvsLow_PFKFB4, fig.width=10, fig.height=7, warning=FALSE}
plotVln(SeuratObject = WT, gene="Pfkfb4", meta=NULL,
	stats=RES,
	vlnsplit = TRUE, fontTXT,  nCell.y=-0.1, pt.alpha=0.4) +
      guides(fill= guide_legend(nrow=1, byrow=TRUE)) 

```

## Conclusion

> PFKFB1/2 is not expressed in plaque cells. 
PFKFB3 is expressed mainly by Mature-DC in atherosclerotic plaques. 
PFKFB4 is expressed mainly by Inflammatory and trem2-foamy Macrophages and MoDC.
However no differences in PFKFB3/4 expression are observed between cells undergoing 
high vs low hypoxia response (stratified by the 3rd quartile of hypoxia response), 
IFNIC-Mac and Mature-DC not tested because low number of cells).

## Save the Seurat Object
```{r seurat_obj}
saveRDS(WT, paste0(DATADIR,"/M.rds"));
```

## SessionInfo
```{r}
sessionInfo()

{                                                                                                                                                                                                           
sink(file=paste0(OUTDIR,"/sessionInfo.txt"))
print(sessionInfo())
sink()
}
```
