---
title: "Analysis invivo PFKFB3 - human plaque cells from Wirka et al 2019"
author: "Javier Perales-Patón - javier.perales@bioquant.uni-heidelberg.de - ORCID: 0000-0003-0780-6683" 
output: github_document
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
## Setup
We define a random seed number for reproducibility, file structure for the output, 
and load essential libraries

### Environment
```{r env}
# Seed number
set.seed(1234)
# Output directory
OUTDIR <- "./Wirka2019/"
if(!dir.exists(OUTDIR)) dir.create(OUTDIR);

# Figures
FIGDIR <- paste0(OUTDIR, "/figures/")
knitr::opts_chunk$set(fig.path=FIGDIR)
knitr::opts_chunk$set(dev=c('png','tiff'))
knitr::opts_chunk$set(dpi=300)
# Data
DATADIR <- paste0(OUTDIR, "/data/")
if(!dir.exists(DATADIR)) dir.create(DATADIR);
```


### Load libraries
```{r}
suppressPackageStartupMessages(require(Seurat))
suppressPackageStartupMessages(require(cowplot))
suppressPackageStartupMessages(require(ggplot2))
suppressPackageStartupMessages(require(purrr))
suppressPackageStartupMessages(require(dplyr))
suppressPackageStartupMessages(require(viper))
suppressPackageStartupMessages(require(rcompanion))
source("../src/graphics.R")
source("../src/seurat_fx.R")
source("../src/wilcox_fx.R")
```

We start with the processed data downloaded from GEO and the metadata information.
```{r}
cnt_file <- gzfile("../data/Wirka_etal/GSE131778_human_coronary_scRNAseq_wirka_et_al_GEO.txt.gz", "rt")
mat <- read.table(cnt_file, sep="\t", header=TRUE, stringsAsFactors = FALSE)
meta <- read.table("../data/Wirka_etal/Wirka_metadata.csv", sep=",", header=TRUE, stringsAsFactors = FALSE)

stopifnot(all(meta$X %in% colnames(mat)))

#NOTE: Standarization : add PFKFB1
if("PFKFB1" %in% rownames(mat)) {
	warning("WARN : Virtually added 'PFKFB1' gene as zero expression\n")
	mat <- rbind(mat, "PFKFB1"=rep(0, ncol(mat)))
}

## Reformat
rownames(meta) <- meta$X
meta <- meta[,colnames(meta)!="X"]
# Add sample
meta$orig <- sapply(rownames(meta), function(z) strsplit(z,split="\\.")[[1]][2])
```

```{r sobj}
S <- CreateSeuratObject(counts=as.matrix(mat), project="Wirka_Human_8n",
			assay = "RNA", min.cells =3, min.features = 500,
			meta.data=meta)
Idents(S) <- S@meta.data$ident
table(Idents(S))
rm(mat, meta)
```

## Normalize data

```{r}
S <- NormalizeData(S)
S <- ScaleData(S)
```

Standarization as our own data
```{r}
# Remove odd clusters named as numbers
S <- S[, grepl("[a-z]",Idents(S))]
S$Annotation.Level.2 <- Idents(S)
# All are WT
WT <- S
# Myeloids
myeloid <- c("Megakaryocyte", "Thrombocyte", "Erythrocyte", "Mast cell",
	     "Basophil", "Neutrophil", "Eosinophil", "Monocyte", "Macrophage")
myeloid_IN <- grep(paste(myeloid, collapse="|"), levels(WT), value=TRUE)

WT$is_myeloid <- sapply(Idents(WT), function(z) ifelse(z%in%myeloid_IN, TRUE, FALSE))
WT$stim <- "WT"
rm(S)
```

```{r reorder}
# Sort them by, 1st Macrophages, then the rest, finally the odd ones
ids <- c(sort(grep("-?Mac", levels(WT), value=TRUE)),
	 sort(grep("-?Mac", levels(WT), value=TRUE, invert=TRUE)))
Idents(WT) <- factor(as.character(Idents(WT)),
		      levels=ids)
# We show the total number of cells
ncol(WT)
# And the sample size of each group
table(Idents(WT))
```


## UMAP
We have subset the data so we do a selection of variables, scaling etc for umap
```{r}
DefaultAssay(WT) <- "RNA"
# WT <- NormalizeData(WT, verbose=FALSE)
WT <- ScaleData(WT, verbose=FALSE)
WT <- FindVariableFeatures(WT, 
			   selection.method = "vst",
			   nfeatures = 2000, verbose = FALSE)
WT <- RunPCA(WT, npcs = 30, verbose = FALSE)
# ElbowPlot(M.i)
WT <- RunUMAP(WT, reduction = "pca", dims = 1:15)
```

```{r Wirka2019_WT_UMAP, fig.width=7, fig.height=5}
p1 <- DimPlot(WT, label=TRUE, repel = TRUE)
print(p1)
```

## Dissection of hypoxia response

### Dorothea focused on HIF1A
We calculate HIF1A transcription factor activities using dorothea. 

```{r dorothea_hif1a, fig.width=6, fig.height=7, dpi=300, warning=FALSE}
df2regulon <- function(df, regulator_name="tf") {
  regulon = df %>% split(.[regulator_name]) %>% map(function(dat) {
    targets = setNames(dat$mor, dat$target)
    likelihood = dat$likelihood
    list(tfmode = targets, likelihood = likelihood)
  })
  return(regulon)
}

regulon.df <- read.table("../data/Prior/dorothea_regulon_human_v1.txt", sep=",", header=TRUE, stringsAsFactors = FALSE)
regul <- df2regulon(df=regulon.df)

# Calculate TF activities
TF <- viper(eset = as.matrix(WT@assays$RNA@data), regulon = regul,
              nes = T, method = "none", minsize = 4,
              eset.filter = F, adaptive.size = F)
 
# Add them as metadata
stopifnot(colnames(WT) == colnames(TF))
WT$HIF1A_activity <- TF["HIF1A",]
rm(TF)
```

```{r HIF1A_strata}
WT$HIF1A_strata <- NA
for(cell in levels(WT)) {
	idx <- WhichCells(WT, idents=cell)
	Q3 <- quantile(WT$HIF1A_activity[idx], probs=0.75)
	WT$HIF1A_strata[idx] <- ifelse(WT$HIF1A_activity[idx] > Q3,
					 "High_HIF1A", "Low_HIF1A")
}
WT$HIF1A_strata <- factor(WT$HIF1A_strata, levels=c("Low_HIF1A", "High_HIF1A"))
```

### PROGENy focused on hypoxia response
We calculate a score of hypoxia response using progeny as well.

```{r vln_progeny_hypoxia, fig.width=6, fig.height=7, dpi=300, warning=FALSE}
### Progeny ####
progeny.mat <- read.table("../data/Prior/progeny_matrix_human_v1.csv",sep=",",header=TRUE)
rownames(progeny.mat) <- progeny.mat$X
progeny.mat <- progeny.mat[which(colnames(progeny.mat)!="X")]
progeny.mat <- as.matrix(progeny.mat)

common <- intersect(rownames(WT), rownames(progeny.mat))
  
prog <- t(as.matrix(WT@assays$RNA@data[common,])) %*% progeny.mat[common,]
rn <- rownames(prog)
prog <- apply(prog,2,scale)
rownames(prog) <- rn
prog <- t(prog)
  
stopifnot(colnames(WT) == colnames(prog))
WT$Hypoxia_response <- prog["Hypoxia",]
rm(common,prog)
```

```{r hypoxia_withinpop_scaled}
WT$Hypoxia_scaled <- NA
for(cell in levels(WT)) {
	idx <- WhichCells(WT, idents=cell)
	WT$Hypoxia_scaled[idx] <- scale(WT$Hypoxia_response[idx])
}
```

```{r hypoxia_strata}
WT$Hypoxia_strata <- NA
for(cell in levels(WT)) {
	idx <- WhichCells(WT, idents=cell)
	Q3 <- quantile(WT$Hypoxia_response[idx], probs=0.75)
	WT$Hypoxia_strata[idx] <- ifelse(WT$Hypoxia_response[idx] > Q3,
					 "High_hypoxia", "Low_hypoxia")
}
WT$Hypoxia_strata<- factor(WT$Hypoxia_strata, levels=c("Low_hypoxia", "High_hypoxia"))
```

## Results
```{r gene}
gene <- "PFKFB3"
genes <- paste0("PFKFB",1:4) 
genes <- intersect(genes, rownames(WT))
```

### PFKFB3
```{r Wirka2019_WT_PFKFB3_FeaturePlots, fig.width=6, fig.height=5}
p2 <- FeaturePlot(WT, features=gene, pt.size = 0.3)
print(p2)
```

```{r Wirka2019_WT_PFKFB3_UMAPFeaturePlots, fig.width=14, fig.height=5}
plot_grid(p1,p2, rel_widths = c(0.55,0.45))
```

```{r Wirka2019_WT_PFKFB3_DotPlots, fig.width=5, fig.height=5}
DotPlot(WT, features=gene) + 
	xlab("") + ylab("Cell type")
```

```{r Wirka2019_WT_PFKFB3_ViolinPlots, fig.width=6, fig.height=5, warning=FALSE}
# VlnPlot(WT, features=gene, pt.size=0.7) + NoLegend()
plotVln(SeuratObject = WT, gene="PFKFB3", meta=NULL,
 	stats=NULL,
	vlnsplit = FALSE, fontTXT,  nCell.y=-0.2, pt.alpha=0.8) +
        NoLegend()
```

### PFKFB isoforms
```{r Wirka2019_WT_PFKFBiso_FeaturePlots, fig.width=7.5, fig.height=6}
p2 <- FeaturePlot(WT, features=genes, pt.size = 0.3)
print(p2)
```

```{r Wirka2019_WT_PFKFBiso_UMAPFeaturePlots, fig.width=14, fig.height=5}
plot_grid(p1,p2, rel_widths = c(0.53,0.47))
```

```{r Wirka2019_WT_PFKFBiso_DotPlots}
DotPlot(WT, features=genes) + 
	xlab("PFKFB isoforms") + ylab("Cell type")
```

```{r Wirka2019_WT_PFKFBiso_ViolinPlots, fig.width=10, fig.height=10, warning=FALSE}
# VlnPlot(WT, features=genes, ncol=length(genes)/2, pt.size=0.7)
plot_grid(plotlist=sapply(genes,function(z) {
		plotVln(SeuratObject = WT, gene=z, meta=NULL,
		  stats=NULL,
		  vlnsplit = FALSE, fontTXT,  nCell.y=-0.1, pt.alpha=0.8) +
			  theme(axis.text.x=element_text(size=16)) + 
			  NoLegend()
		 }, simplify=FALSE), ncol = ceiling(length(genes)/2))
```


### PFKFB3 High Hypoxia/HIF1A

#### Hypoxia
```{r Wirka2019_WThypoxiaHighvsLow_PFKFB3_stats}
WT$stim <- WT$Hypoxia_strata
wPFKFB3_stats <- t(sapply(levels(WT), function(cell) {
				   cellIds <- WhichCells(WT, idents=cell)
				   # The test
				   wilcox.test_stats(xy=WT@assays$RNA@data["PFKFB3", cellIds],
						     gr=WT$stim[cellIds])
			   }))
wPFKFB3_stats <- as.data.frame(wPFKFB3_stats)
wPFKFB3_stats$adjpval <- p.adjust(wPFKFB3_stats$pvalue, method="fdr")
wPFKFB3_stats$significance <- tagSignif(wPFKFB3_stats$adjpval)
print(wPFKFB3_stats)

RES <- wPFKFB3_stats[, c("r", "significance")]
RES$r <- format(round(RES$r, 2), nsmall=2)
colnames(RES)[ncol(RES)] <- ""
```


```{r Wirka2019_WThypoxiaHighvsLow_PFKFB3, fig.width=10, fig.height=7, warning=FALSE}
plotVln(SeuratObject = WT, gene="PFKFB3", meta=NULL,
	stats=RES,
	vlnsplit = TRUE, fontTXT,  nCell.y=-0.1, pt.alpha=0.4) +
      guides(fill= guide_legend(nrow=1, byrow=TRUE)) 

```

#### HIF1A

```{r Wirka2019_WTHIF1AHighvsLow_PFKFB3_stats}
WT$stim <- WT$HIF1A_strata
wPFKFB3_stats <- t(sapply(levels(WT), function(cell) {
				   cellIds <- WhichCells(WT, idents=cell)
				   # The test
				   wilcox.test_stats(xy=WT@assays$RNA@data["PFKFB3", cellIds],
						     gr=WT$stim[cellIds])
			   }))
wPFKFB3_stats <- as.data.frame(wPFKFB3_stats)
wPFKFB3_stats$adjpval <- p.adjust(wPFKFB3_stats$pvalue, method="fdr")
wPFKFB3_stats$significance <- tagSignif(wPFKFB3_stats$adjpval)
print(wPFKFB3_stats)

RES <- wPFKFB3_stats[, c("r", "significance")]
RES$r <- format(round(RES$r, 2), nsmall=2)
colnames(RES)[ncol(RES)] <- ""
```

```{r Wirka2019_WTHIF1AHighvsLow_PFKFB3, fig.width=10, fig.height=7, warning=FALSE}
plotVln(SeuratObject = WT, gene="PFKFB3", meta=NULL,
	stats=RES,
	vlnsplit = TRUE, fontTXT,  nCell.y=-0.1, pt.alpha=0.4) +
      guides(fill= guide_legend(nrow=1, byrow=TRUE))

```

### PFKFB4 High Hypoxia/HIF1A


#### Hypoxia
```{r Wirka2019_WThypoxiaHighvsLow_PFKFB4_stats}
WT$stim <- WT$Hypoxia_strata
wPFKFB4_stats <- t(sapply(levels(WT), function(cell) {
				   cellIds <- WhichCells(WT, idents=cell)
				   # The test
				   wilcox.test_stats(xy=WT@assays$RNA@data["PFKFB4", cellIds],
						     gr=WT$stim[cellIds])
			   }))
wPFKFB4_stats <- as.data.frame(wPFKFB4_stats)
wPFKFB4_stats$adjpval <- p.adjust(wPFKFB4_stats$pvalue, method="fdr")
wPFKFB4_stats$significance <- tagSignif(wPFKFB4_stats$adjpval)
print(wPFKFB4_stats)

RES <- wPFKFB4_stats[, c("r", "significance")]
RES$r <- format(round(RES$r, 2), nsmall=2)
colnames(RES)[ncol(RES)] <- ""
```


```{r Wirka2019_WThypoxiaHighvsLow_PFKFB4, fig.width=10, fig.height=7, warning=FALSE}
plotVln(SeuratObject = WT, gene="PFKFB4", meta=NULL,
	stats=RES,
	vlnsplit = TRUE, fontTXT,  nCell.y=-0.1, pt.alpha=0.4) +
      guides(fill= guide_legend(nrow=1, byrow=TRUE)) 

```

#### HIF1A
```{r Wirka2019_WTHIF1AHighvsLow_PFKFB4_stats}
WT$stim <- WT$HIF1A_strata
wPFKFB4_stats <- t(sapply(levels(WT), function(cell) {
				   cellIds <- WhichCells(WT, idents=cell)
				   # The test
				   wilcox.test_stats(xy=WT@assays$RNA@data["PFKFB4", cellIds],
						     gr=WT$stim[cellIds])
			   }))
wPFKFB4_stats <- as.data.frame(wPFKFB4_stats)
wPFKFB4_stats$adjpval <- p.adjust(wPFKFB4_stats$pvalue, method="fdr")
wPFKFB4_stats$significance <- tagSignif(wPFKFB4_stats$adjpval)
print(wPFKFB4_stats)

RES <- wPFKFB4_stats[, c("r", "significance")]
RES$r <- format(round(RES$r, 2), nsmall=2)
colnames(RES)[ncol(RES)] <- ""
```


```{r Wirka2019_WTHIF1AHighvsLow_PFKFB4, fig.width=10, fig.height=7, warning=FALSE}
plotVln(SeuratObject = WT, gene="PFKFB4", meta=NULL,
	stats=RES,
	vlnsplit = TRUE, fontTXT,  nCell.y=-0.1, pt.alpha=0.4) +
      guides(fill= guide_legend(nrow=1, byrow=TRUE)) 

```

## Save the Seurat Object
```{r seurat_obj}
saveRDS(WT, paste0(DATADIR,"/M.rds"));
```

## SessionInfo
```{r}
sessionInfo()

{                                                                                                                                                                                                           
sink(file=paste0(OUTDIR,"/sessionInfo.txt"))
print(sessionInfo())
sink()
}
```
