---
title: "Exploratory analysis on conventional bulk RNA-seq from PHD2 in atherogenesis"
author: "Javier Perales-Paton - javier.perales@bioquant.uni-heidelberg.de" 
---

Transcriptome profiling with conventional RNAseq was performed in three replicates of macrophages 
carrying PHD2 knock-out (with wildtype) and fibroblasts co-cultured with supernadant of these 
macrophages (as compared to supernadant of the wiltype genotype). Herein we perform an exploratory
analysis on these settings.

## Setting-up environment
The environment will be set with a random seed number for reproducibility and an output folder
for processed data and figures.

### set env
```{r env}
options(stringsAsFactors = FALSE)
# Seed number
set.seed(1234)
# Output directory
OUTDIR <- "./output/00_exploratory"
if(!dir.exists(OUTDIR)) dir.create(OUTDIR);
```

### Load libraries
```{r}
library(edgeR)
library(limma)
library(fgsea)
library(GSEABase)
library(dendextend)
```

## Load data and normalize
```{r}
### 1 Load data
cnt <- read.table("../data/bulk/MST_MC_N_fibroblasts_PHD2-rawdata.txt", sep="\t", header = TRUE, check.names = FALSE)
rownames(cnt) <- cnt[,1]
cnt <- cnt[,-1]

targets <- read.table("../data/bulk/MST_MC_N_fibroblasts_PHD2-metadata.txt", sep="\t",header=TRUE, colClasses = "character")
targets$Sample <- gsub(" ","_",targets$Sample)

# Rename samples
stopifnot(all(colnames(cnt)==targets$FastQ))
colnames(cnt) <- targets$Sample

# Create metadata
stopifnot(all(colnames(cnt)==targets$Sample))
gr <- targets$Group
gt <- factor(gsub("^(3T3|MΦ)_", "", targets$Group), levels=c(c("PHD2_WT", "PHD2_KO")))
cell <- factor(unlist(sapply(targets$Group, function(z) strsplit(z, split="_")[[1]][1])))
repl <- factor(targets$Replicate)

## 2 Create DGElist
y <- DGEList(counts = cnt, group = gr, genes = rownames(cnt))
y <- calcNormFactors(y, method = "TMM")
```

## Diagnostics
### The library size
```{r}
par(mar=c(10,8,4,4)); 
barplot(y$samples$lib.size,names.arg = rownames(y$samples), las=2)
```

### Principal Component analysis
First we start with the definition of ad-hoc handle functions.
```{r}
# Define some functions
runPCA <- function(mat) {
  pca <- prcomp(t(mat))
  
  return(pca)
}

plotPCA <- function(pca, dims, pchs, cols, labels=NULL) {
  importance <- summary(pca)$importance[,dims]
  PCscores <- data.frame(pca$x)[,dims]
  
  plot(PCscores,las=1,
       pch=pchs,bg=cols,
       xlab=paste0(dims[1]," : ",format(round(importance[2,dims[1]]*100,digits = 2),nsmall = 2),"% var expl."),
       ylab=paste0(dims[2]," : ",format(round(importance[2,dims[2]]*100,digits = 2),nsmall = 2),"% var expl."))
  if(!is.null(labels)) {
    par(xpd=TRUE)
    text(x=PCscores[,dims[1]], y=PCscores[,dims[2]], labels=labels, pos=1)
  }
}

```

Then we run PCA and visualize it
```{r pca, dpi=300, fig.width=7, fig.height=5, fig.path="./output/00_exploratory/", dev=c('png', 'tiff')}

# Run PCA
pca <- runPCA(mat = cpm(y, log = TRUE))

# Visualize
 #plot(PC2 ~ PC1, data=data.frame(pca$x),pch=c(21,23)[as.integer(cell)],bg=as.integer(gt))
 #par(xpd=TRUE); text(x = pca$x[,1], y=pca$x[,2], labels = targets$Replicate, pos = 2)

par(mar=c(4,4,4,12), xpd=TRUE)
plotPCA(pca=runPCA(mat = cpm(y, log = TRUE)),
        dims=c("PC1","PC2"),
        pchs=c(21,23)[as.integer(cell)], cols=as.integer(gt))
legend("right",legend = levels(cell), pch=c(21,23), 
       inset = c(-0.3,0), title = "Cell type")
legend("topright",legend = levels(gt), pch=c(22),pt.bg = c("black","red"), 
       inset = c(-0.4,0), title = "Perturbation")

# Which is
PC1_loadings <- sort(abs(pca$rotation[,"PC1"]),decreasing = TRUE)
GSC <- getGmt("../data/MSigDB/h.all.v6.2.symbols.gmt")
res <- fgsea(pathways = geneIds(GSC), stats = setNames(PC1_loadings, toupper(names(PC1_loadings))), nperm = 1000)
res <- res[order(res$padj,decreasing = FALSE),]

options(digits=2)
print(res[, 1:5])
```

We could conclude that PHD2 perturbation is the main factor of variability in the experiment.
Moreover, It seems Hypoxia response is involved in this separation. 
This makes sense given the knowledge on the role of PHD2.

### Hierchical clustering

```{r hc, dpi=300, fig.width=6, fig.height=7, fig.path="./output/00_exploratory/", dev=c('png','tiff')}
cpm_norm <- cpm(y,log=TRUE)
rowise_var <- apply(cpm_norm,1,var)
genes_qt75 <- names(rowise_var[rowise_var>= quantile(rowise_var, 0.90)])
cpm_transform <- sweep(cpm_norm, 1, apply(cpm_norm, 1, median, na.rm=T))

d <- dist(t(cpm_transform[genes_qt75, ]))
hc <- hclust(d, method = "complete")
dend <- as.dendrogram(hc)

# Color the branches based on the clusters:
dend <- color_branches(dend, k=2) 
dend <- assign_values_to_branches_edgePar(dend=dend, value=c(1.5,1.5), edgePar="lwd")

# plot(dend, las=1, ylab="height")

# Manually match the labels, as much as possible, to the real classification of the flowers:
labels_colors(dend) <- as.integer(factor(setNames(gt, 
						  targets$Sample)[labels(dend)]))

par(mar=c(8,5,4,4))
plot(dend, las=1, ylab="", cex.axis=1.5)
title(ylab="height", cex.lab=1.5, line=+3.7)
```

