---
title: "Inference of TF activities in Mieloid PHD2 deficiency in atherogenesis"
author: "Javier Perales-Paton - javier.perales@bioquant.uni-heidelberg.de"
output: github_document
---

We will use the ranking of differentially expressed genes by the moderated-t statistics 
to estimate the pathways activities using PROGENy. Then, we will focused on Hypoxia response 
in macrophages upon PHD2 knock-out given two facts:
* Exploratory data analysis (EDA) has shown that hypoxia is driving the major pattern of 
  transcriptional heterogeneity in the dataset, separating the two conditions (PHD2cKO vs WT).
* Actually the literature has described that PHD2 impair leads to Hif1a stabilization, which
  triggers hypoxia response.


## Set-up environment
Define randome seed and output folder.
```{r env}
# Seed number
set.seed(1234)
# Output directory
OUTDIR <- "./02_pathway_output/"
if(!dir.exists(OUTDIR)) dir.create(OUTDIR);

# Figures
FIGDIR <- paste0(OUTDIR, "/figures/")
knitr::opts_chunk$set(fig.path=FIGDIR)
knitr::opts_chunk$set(dev=c('png','tiff'))
# Data
DATADIR <- paste0(OUTDIR, "/data/")
if(!dir.exists(DATADIR)) dir.create(DATADIR);
```

## Load libraries
Essential R libraries.
```{r}
library(limma)
library(progeny)
```

## Load data
We use the output data from previous step (01). In particular,
the eBay object from limma which contains the contrasts and statistics.
```{r}
eBay <- readRDS("./01_DGE_output/data/eBay.rds")
```

## PROGENy analysis

```{r}
source("../src/runProgenyFast.R") # Ack. to A. Dugourd (c under GPL license)
progeny.mat <- read.table("../data/Prior/progeny_matrix_mouse_v1.txt",sep=",",header=TRUE)
df <- data.frame(ID=rownames(eBay$t),
                 Fib_PHD2=eBay$t[,"Fib_PHD2"],
                 MC_PHD2=eBay$t[,"MC_PHD2"])

set.seed(1234)
progeny.res <- runProgenyFast(df, progeny.mat)
pvals <- apply(progeny.res, 2, function(z) pnorm(z))
pvals <- apply(pvals, c(1,2), function(pval) ifelse(pval > 0.5, (1-pval)*2, pval*2))
colnames(pvals) <- colnames(progeny.res)

for(tag in names(progeny.res)) {
    # Make a simple table with the outcome
    progeny.cont <- cbind(Activity=progeny.res[,tag],Pvalue=pvals[,tag])
    # Show in stdout
    cat(paste("Pathway activity from",tag,"\n"), file=stdout())
    print(progeny.cont)
    # Save if as a supplementary table in a file
  write.table(progeny.cont,
              file = paste0(DATADIR,"/",tag,"_progeny.tsv"),
              sep="\t", col.names = NA, row.names = TRUE)
}
```
It seems that hypoxia is highly active in Macrophages upon PHD2-KO
(z-score=14.85, pval\<0.0001). Thus we make a visualization of this
activation for publication.

```{r macrophages_hypoxia, fig.width=8, fig.height=6, dpi=300, warning=FALSE}
source("../src/progenyScores.R")
source("../src/graphics.R")
progeny_scatter.out <- progenyScatter(df, progeny.mat, fontfamily=fontTXT)

plot(progeny_scatter.out[[2]]$Hypoxia)
```

## Session info

```{r}
sessionInfo()

{                                                                                                                                                                                                           
sink(file=paste0(OUTDIR,"/sessionInfo.txt"))
print(sessionInfo())
sink()
}
```

