---
title: "Inference of Transcription Factor activity in Mieloid PHD2 deficiency in atherogenesis"
author: "Javier Perales-Patón - javier.perales@bioquant.uni-heidelberg.de - ORCID: 0000-0003-0780-6683" 
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Herein we perform a Transcription Factor (TF) analysis on the role of Myeloid-derived PHD2 in 
atherogenesis. For that, we use normalized data from the differential gene expression analysis 
and estimate TF activities using DoRothEA mouse regulons. See main README file for details.

## Setup environment
The environment will be set with a random seed number for reproducibility and an output folder
for processed data and figures.

### set env
```{r env}
options(stringsAsFactors = FALSE)
# Seed number
set.seed(1234)
# Output directory
OUTDIR <- "./03_TF_output/"
if(!dir.exists(OUTDIR)) dir.create(OUTDIR);

# Figures
FIGDIR <- paste0(OUTDIR, "/figures/")
knitr::opts_chunk$set(fig.path=FIGDIR)
knitr::opts_chunk$set(dev=c('png','tiff'))
# Data
DATADIR <- paste0(OUTDIR, "/data/")
if(!dir.exists(DATADIR)) dir.create(DATADIR);
```

### Load libraries
Essential libraries for R analysis.
```{r load_libs}
library(limma)
library(viper)
library(purrr)
```

## Load data
We use the output data from DGE step (01). In particular,
the limma objects which contains gene expression, statistics and 
design of the contrasts.
```{r}
v <- readRDS("./01_DGE_output/data/v.rds")
eBay <- readRDS("./01_DGE_output/data/eBay.rds")
```

## Transcription factor analysis using DoRothEA
We read the mouse regulons for TF analysis.
```{r}
# Function to transform a data.frame to viper's regulon 
# format using purrr. Ack. to CH. Holland.
df2regulon <- function(df, regulator_name="tf") {
  regulon = df %>% split(.[regulator_name]) %>% map(function(dat) {
    targets = setNames(dat$mor, dat$target)
    likelihood = dat$likelihood
    list(tfmode = targets, likelihood = likelihood)
  })
  return(regulon)
}

# Read dorothea mouse TF regulons
regulon.df <- read.table("../data/Prior/dorothea_regulon_mouse_v1.csv", 
			 sep=",", header=TRUE, stringsAsFactors = FALSE)
# We shrink the analysis to those TF regulons with high-to-intermediate confidence
regulon.df <- subset(regulon.df, confidence %in% c("A","B","C")) # E are discarded
# Transform data.frame to viper regulon format
regul <- df2regulon(df=regulon.df)
```

The TF activities are estimated using `viper`. Similarly to gene-level differential expression, 
contrasts performed using Empirical Bayes method implemented in `limma`
```{r}
TF = viper(eset = v$E, 
	   regulon = regul,
           nes = T, minsize = 4,
           eset.filter = F, 
	   adaptive.size = F, 
	   verbose=FALSE)

TF.fit <- lmFit(TF, v$design)
TF.cont.mat <- makeContrasts("Fib_PHD2"=Fib_PHD2_KO - Fib_PHD2_WT,
                          "MC_PHD2"=MC_PHD2_KO - MC_PHD2_WT,
                          levels=v$design)
TF.fit <- eBayes(TF.fit)
TF.fit2 <- contrasts.fit(TF.fit, TF.cont.mat)
TF.eBay <- eBayes(TF.fit2)
```


## Broad statistics on diff TFs
Just check how many TFs are differentially active
```{r}
diffTF_cnt <- apply(decideTests(TF.eBay),2, table)
print(diffTF_cnt)
```

## Diagnostics and save results
Write tables as supplementary data for the records and further interpretation.
```{r diagnostics_diffTF}
for(cont in colnames(TF.cont.mat)) {
  cat(paste0("Registering differential expression for ",cont,"\n"),
	file=stdout())
  topTab <- topTable(TF.eBay, coef=cont, number = Inf)
  # DEGs table
  write.table(topTab, file=paste0(DATADIR,cont,"_diffTF.tsv"),sep="\t",
                                  row.names=TRUE, col.names=NA, quote=FALSE)
  # Volcano plot
  plot(TF.eBay$coefficients[,cont], -log10(TF.eBay$p.value[,cont]),
       xlab="log2-fold-change", ylab="-log10(pvalue)", cex=0.7)
  # Histogram of p-vals
  hist(TF.eBay$p.value[,cont])
}
```
Histogram of p-values looks good, and volcano plots as well. Certainly Fibroblasts presents 
higher proportion of TF dysregulated from the medium conditioned with Macrophages PHD2cKO.

## Hypothesis-driven test of Hif1a up-regulation in Macrophages PHD2cKO
Based on previous pathway analysis and literature, it would be expected that actually 
Hif1a transcription factors presents high activity upon PHD2 dysfunction. As it is confirmed 
below:

```{r Hif1a_UP_confirmation}
MC_PHD2cKO_Hif1a <- topTable(TF.eBay, coef="MC_PHD2", number=Inf)["Hif1a",]
```

We could conclude that Hif1a is `r 2^MC_PHD2cKO_Hif1a$logFC`-fold times more active 
in macrophages upon PHD2-KO (FDR=`r MC_PHD2cKO_Hif1a$adj.P.Val`).

## Save data
```{r}
saveRDS(TF, file=paste0(DATADIR,"/v.rds"))
saveRDS(TF.eBay, file=paste0(DATADIR,"/eBay.rds"))
```

## Session info

```{r}
sessionInfo()

{                                                                                                                                                                                                           
sink(file=paste0(OUTDIR,"/sessionInfo.txt"))
print(sessionInfo())
sink()
}
```
